---
title: "Late-onset toxicity with the TITE-BOIN design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Late-onset toxicity with the TITE-BOIN design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(simFastBOIN)
```

## Why a time-to-event design

The BOIN design needs every patient's outcome before the next cohort is dosed.
When toxicity can appear late in a long assessment window, or patients arrive
faster than they can be assessed, that means repeatedly suspending accrual. The
time-to-event BOIN (TITE-BOIN) design treats new patients while earlier ones are
still being followed, and uses the follow-up time of these pending patients. Two
ways of doing so are available through `method`: the single mean imputation of
Yuan et al. (2018), the default, and the effective sample size of Lin and Yuan
(2020).

## The decision table

The decision table now depends on three counts, the patients treated, the DLTs
observed and the patients pending, and in some states on the standardized total
follow-up time (STFT) of the pending patients, their total follow-up divided by
the length of the window. The table for a target of 0.2 is Table 1 of Yuan et
al. (2018), which the package reproduces entry by entry.

```{r}
tite <- tite_boin_decision_table(target = 0.2, max_n = 15)
print(tite, cohort_size = 3)
```

The table is a data frame, so a single state can be looked up. With nine
patients, one DLT and four pending, the dose is escalated once STFT reaches
2.15. In the trial example of Yuan et al. (2018) the four pending patients had
been followed for 75, 60, 45 and 30 days of a 90 day window, an STFT of 2.33,
and the dose was escalated.

```{r}
tite[tite$n == 9 & tite$n_tox == 1 & tite$n_pending == 4, ]
```

The same table as a figure, here for nine patients. Each cell is a pair of
pending patients and DLTs; a cell whose decision depends on the follow-up shows
the boundaries, the one after `>=` for escalation and the one after `<=` for
de-escalation.

```{r tite-plot, fig.width = 7, fig.height = 6, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot(tite, n = 9)
```

Under the effective sample size the boundaries refer to ESS, the number of
patients who have completed the assessment plus STFT, and escalation waits until
two patients at the dose have completed it.

```{r}
print(tite_boin_decision_table(target = 0.3, max_n = 9, method = "ess"),
      cohort_size = 3)
```

## Operating characteristics

`sim_tite_boin()` simulates the design with patients arriving over time. It
needs the length of the assessment window and the accrual rate, in the same unit
of time, and reports the trial duration and the suspensions of accrual in
addition to the summaries of `sim_boin()`.

```{r}
oc_tite <- sim_tite_boin(
  target = 0.30,
  p_true = c(0.05, 0.15, 0.30, 0.45, 0.60),
  n_cohort = 10,
  cohort_size = 3,
  window = 3,
  accrual_rate = 2,
  n_trials = 2000,
  seed = 123
)

oc_tite
```

Faster accrual shortens the trial, at the price of more frequent suspensions.

```{r}
by_rate <- sapply(c(0.5, 1, 2, 4), function(rate) {
  oc <- sim_tite_boin(target = 0.30, p_true = c(0.05, 0.15, 0.30, 0.45, 0.60),
                      n_cohort = 10, cohort_size = 3, window = 3,
                      accrual_rate = rate, n_trials = 1000, seed = 1)
  c(accrual_rate = rate, selected_dose_3 = oc$sel_percent[["DL3"]],
    duration = oc$duration_mean, suspensions = oc$avg_n_suspensions)
})
round(t(by_rate), 2)
```

## Agreement with the BOIN design

When no patient is ever pending at a decision, the time-to-event design is the
BOIN design, and the simulation engine reproduces `sim_boin()` exactly. With
arrivals one unit of time apart and a window of half a unit, every patient has
completed the assessment before the next one arrives.

```{r}
no_pending <- sim_tite_boin(target = 0.30,
                            p_true = c(0.05, 0.15, 0.30, 0.45, 0.60),
                            n_cohort = 10, cohort_size = 3, window = 0.5,
                            accrual_rate = 1, accrual = "fixed",
                            n_trials = 1000, seed = 7)
plain_boin <- sim_boin(target = 0.30, p_true = c(0.05, 0.15, 0.30, 0.45, 0.60),
                       n_cohort = 10, cohort_size = 3, n_trials = 1000,
                       seed = 7)

identical(no_pending$sel_percent, plain_boin$sel_percent)
identical(no_pending$n_pts_dose, plain_boin$n_pts_dose)
```

## References

Liu S. and Yuan, Y. (2015). Bayesian Optimal Interval Designs for Phase I
Clinical Trials. *Journal of the Royal Statistical Society: Series C*, 64,
507-523.

Yuan, Y., Lin, R., Li, D., Nie, L. and Warren, K. E. (2018). Time-to-Event
Bayesian Optimal Interval Design to Accelerate Phase I Trials. *Clinical Cancer
Research*, 24(20), 4921-4930.

Lin, R. and Yuan, Y. (2020). Time-to-Event Model-Assisted Designs for
Dose-Finding Trials with Delayed Toxicity. *Biostatistics*, 21(4), 807-824.
