---
title: "Backfilling with the BF-BOIN and BE-BOIN designs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Backfilling with the BF-BOIN and BE-BOIN designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(simFastBOIN)
```

## Why backfill

In the BOIN design the next cohort is dosed only once the current one has
completed the DLT assessment, and patients who become eligible in the meantime
are usually turned away. Backfilling treats them at a lower dose that has
already been cleared, so that more patients contribute data on safety and
activity at doses that may later be carried forward, while the dose escalation
goes on as before.

Two designs are provided. BF-BOIN (Zhao et al., 2024) keeps the BOIN decision
table and fully staggered cohorts. BE-BOIN (Chen et al., 2026) combines
backfilling with the time-to-event BOIN design, so that the escalation itself
decides while some patients are still pending. In both, the package selects the
MTD from the data of all patients, backfilled ones included. The second stage of
Chen et al. (2026), which randomizes patients between doses to choose the
optimal biological dose, is not part of the package.

## The rules of BF-BOIN

The trial has two parts that run side by side.

* **Dose escalation.** Cohorts are fully staggered. When every patient of the
  current cohort has completed the assessment, the next dose follows the BOIN
  decision table applied to the completed patients at the current dose,
  backfilled ones included. Elimination is checked at the same moments, at
  every dose.
* **Backfilling.** A patient who arrives while the escalation waits is treated
  at the highest dose below the current one that has not been eliminated, has
  fewer than `n_cap` patients (12 by default) and has a response observed at or
  below it. A dose is closed when its own DLT rate exceeds the de-escalation
  boundary and so does the rate pooled with the next higher dose; the doses
  above it close with it.
* **Conflicts.** When the data of a backfilled dose call for a more cautious
  decision than those of the current dose (Table 2 of Zhao et al., 2024), the
  decision is taken from the data pooled from that dose up to the current one.

The trial ends when the escalation has treated `n_cohort` cohorts, when
`n_earlystop` patients, backfilled ones included, have been treated at the
current dose and the decision is to stay, or when the lowest dose is
eliminated.

## Operating characteristics of BF-BOIN

Scenario 3 of Zhao et al. (2024) has the MTD at dose 3 and responses that rise
with the dose. The design is that of the article: a target of 0.25, ten
cohorts of three, `n_earlystop = 9`, one DLT out of three treated as a stay,
three patients a month and an assessment window of one month.

```{r}
zhao_design <- list(target = 0.25, n_cohort = 10, cohort_size = 3,
                    window = 1, accrual_rate = 3, n_earlystop = 9,
                    stay_on_1_of_3 = TRUE, accrual = "uniform",
                    no_slot = "leave", n_trials = 2000, seed = 123)
p_true <- c(0.04, 0.12, 0.25, 0.43, 0.63)
p_resp <- c(0.10, 0.20, 0.30, 0.45, 0.58)

oc_bf <- do.call(sim_bf_boin,
                 c(zhao_design, list(p_true = p_true, p_resp = p_resp)))
oc_bf
```

Without any response no dose is opened for backfilling and the design is the
BOIN design. The table below sets the two runs next to the values that Zhao et
al. (2024) report in their Table 4: the percentage of trials selecting the MTD
and a dose above it, the average number of patients below, at and above the MTD
and in all, and the average duration in months.

```{r}
oc_boin <- do.call(sim_bf_boin,
                   c(zhao_design, list(p_true = p_true, p_resp = rep(0, 5))))

zhao_metrics <- function(oc) {
  c(oc$sel_percent[[3]], sum(oc$sel_percent[4:5]),
    sum(oc$n_pts_dose[1:2]), oc$n_pts_dose[[3]], sum(oc$n_pts_dose[4:5]),
    oc$total_n_pts, oc$duration_mean)
}
comparison <- rbind(
  "BF-BOIN, Table 4" = c(57.6, 9.7, 16.9, 9.8, 4.2, 30.9, 14.3),
  "BF-BOIN, simulated" = zhao_metrics(oc_bf),
  "BOIN, Table 4" = c(56.7, 12.3, 11.4, 8.0, 4.6, 23.9, 14.3),
  "BOIN, simulated" = zhao_metrics(oc_boin)
)
colnames(comparison) <- c("MTD (%)", "Above (%)", "n below", "n at",
                          "n above", "n total", "Months")
round(comparison, 1)
```

With 2,000 trials the simulated values carry a simulation error of about one
point for the percentages. The tests of the package compare all twelve
scenarios of Table 4 with 10,000 trials each; over 20 seeds the largest gaps
were 2.1 points for the selection, 0.4 patients and 0.17 months. The script
`inst/validation/compare-with-Zhao2024.R` repeats the comparison.

The response profile decides which doses are backfilled. Scenarios 2 and 6 of
the article share the DLT probabilities but not the response probabilities.

```{r}
scenarios <- list(
  list(name = "Scenario 2", p_true = c(0.12, 0.25, 0.42, 0.49, 0.55),
       p_resp = c(0.20, 0.30, 0.40, 0.50, 0.60)),
  list(name = "Scenario 6", p_true = c(0.12, 0.25, 0.42, 0.49, 0.55),
       p_resp = c(0.30, 0.35, 0.36, 0.36, 0.36))
)
zhao_multi <- zhao_design[names(zhao_design) != "n_trials"]
do.call(sim_bf_boin_multi,
        c(zhao_multi, list(scenarios = scenarios, n_trials = 1000)))
```

## Points the articles leave open

The articles do not settle every detail of a simulated trial. The package
makes the following choices, and offers an argument where more than one
reading is reasonable.

* **A patient with nowhere to go.** When the escalation waits and no dose is
  open for backfilling, `no_slot = "wait"` (the default) lets the patient wait
  until one can be treated, as in `sim_tite_boin()`, and `no_slot = "leave"`
  turns the patient away. Zhao et al. (2024) state a Poisson process of
  arrivals, but their durations, for BOIN as well as for BF-BOIN, agree with
  uniform times between arrivals (`accrual = "uniform"`) and patients turned
  away, which is why the comparison above uses these settings.
* **The end of backfilling.** Backfilling stops when the last patient of the
  escalation is enrolled, and every enrolled patient is then followed to the
  end of the assessment.
* **Pending backfill patients.** The escalation of BF-BOIN uses only the
  patients who have completed the assessment.
* **Responses.** Neither article states how the time to response is
  distributed or how a response is related to a DLT. The time to response is
  Weibull, with `resp_late_fraction` of the responses in the second half of
  `resp_window`, and the DLT and the response are linked by a Gaussian copula
  with correlation `resp_cor`, as in Takeda et al. (2025), who set it to 0.2.
  The default here is 0, independence.
* **Which conflicting dose anchors the pooled estimate.** With several
  conflicting backfilled doses, Zhao et al. (2024) pool from the highest of
  them and Chen et al. (2026) from the lowest. `conflict_dose` chooses, with
  the default of each article.

## BE-BOIN

BE-BOIN decides the dose for a cohort when its first patient arrives, with the
time-to-event BOIN rule, so the escalation waits only when the rule asks it to.
Two rules of Chen et al. (2025) decide when that is: rule 1 suspends accrual
when fewer than 51 percent of the patients at the current dose have completed
the assessment (`max_pending_ratio = 0.49`), and rule 2 allows escalation only
when every pending patient at the current dose has been followed for at least
a quarter of the window (`min_follow_up = 0.25`). The decision table with
both rules is Table A1 of the supplementary materials of Chen et al. (2026),
which `tite_boin_decision_table()` reproduces entry by entry; an escalation
that needs rule 2 is marked `MF >= 0.25`.

```{r}
be_table <- tite_boin_decision_table(target = 0.25, max_n = 9,
                                     max_pending_ratio = 0.49,
                                     min_follow_up = 0.25)
print(be_table, cohort_size = 3)
```

Every estimate of BE-BOIN, for the escalation, for opening and closing
backfill doses and for conflicts, imputes the pending patients with the single
mean imputation of TITE-BOIN. Patients who arrive while the escalation is
suspended are backfilled with the rules of BF-BOIN. With a window of three
months and two patients a month, the two designs can be compared on the same
scenario; Chen et al. (2026) report shorter trials for BE-BOIN.

```{r}
slow_design <- list(target = 0.25, p_true = p_true, p_resp = p_resp,
                    n_cohort = 10, cohort_size = 3, window = 3,
                    accrual_rate = 2, n_earlystop = 9, n_trials = 2000,
                    seed = 123)
oc_be <- do.call(sim_be_boin, slow_design)
oc_be

oc_bf_slow <- do.call(sim_bf_boin, slow_design)
round(rbind(
  "BF-BOIN" = c(MTD = oc_bf_slow$sel_percent[[3]],
                patients = oc_bf_slow$total_n_pts,
                backfilled = oc_bf_slow$total_n_bf,
                months = oc_bf_slow$duration_mean),
  "BE-BOIN" = c(MTD = oc_be$sel_percent[[3]],
                patients = oc_be$total_n_pts,
                backfilled = oc_be$total_n_bf,
                months = oc_be$duration_mean)
), 1)
```

## Agreement with the BOIN and TITE-BOIN designs

The DLT variates of the escalation are drawn from R's stream exactly as in
`sim_boin()`, and arrivals, backfill patients and responses have generators of
their own. When no dose is ever opened for backfilling, the trials are
therefore those of `sim_boin()` with the same seed, and those of
`sim_tite_boin()` for BE-BOIN with `no_slot = "wait"`.

```{r}
no_backfill <- sim_bf_boin(target = 0.25, p_true = p_true,
                           p_resp = rep(0, 5), n_cohort = 10,
                           cohort_size = 3, window = 1, accrual_rate = 3,
                           n_trials = 1000, seed = 7)
plain_boin <- sim_boin(target = 0.25, p_true = p_true, n_cohort = 10,
                       cohort_size = 3, n_trials = 1000, seed = 7)

all.equal(no_backfill$sel_percent, plain_boin$sel_percent)
all.equal(no_backfill$n_pts_dose, plain_boin$n_pts_dose)
```

## How the implementation is checked

* The operating characteristics of Table 4 of Zhao et al. (2024), twelve
  scenarios of BF-BOIN and five of BOIN, are reproduced within simulation
  error, durations included.
* The trial example of Figure 1 of Zhao et al. (2024) and that of Section S2 of
  the supplementary materials of Chen et al. (2026) are replayed patient by
  patient, with the random numbers scripted to give the outcomes of the
  articles.
* The decision tables of Chen et al. (2025) and of the supplementary materials
  of Chen et al. (2026) are reproduced entry by entry.
* The simulated trials agree exactly, trial by trial, with a separate
  implementation written in Python and run on a replica of R's random number
  stream, over 2,700 trials of both designs and several settings.
* Without backfilling the trials are identical to those of the BOIN and
  TITE-BOIN engines, as shown above.

Chen et al. (2026) give their scenarios only as figures, so their operating
characteristics cannot be reproduced number for number.

## References

Zhao, Y., Yuan, Y., Korn, E. L. and Freidlin, B. (2024). Backfilling Patients
in Phase I Dose-Escalation Trials Using Bayesian Optimal Interval Design
(BOIN). *Clinical Cancer Research*, 30(4), 673-679.

Chen, K., Zhao, Y., Takeda, K. and Yuan, Y. (2026). BE-BOIN: A Dose
Optimization Design Accommodating Backfill and Late-Onset Toxicity.
*Therapeutic Innovation and Regulatory Science*.

Chen, K., Chen, T.-Y., Zhang, Y., Lin, R. and Yuan, Y. (2025). Practical
Considerations for Using the TITE-BOIN Design to Handle Late-Onset Toxicity or
Fast Accrual in Phase I Trials. *Clinical Cancer Research*, 31(13), 2573-2580.

Takeda, K., Zhu, J. and Hirakawa, A. (2025). BF-BOIN-ET: A Backfill Bayesian
Optimal Interval Design Using Efficacy and Toxicity Outcomes for Dose
Optimization. *Pharmaceutical Statistics*, 24(2), e2470.

Liu, S. and Yuan, Y. (2015). Bayesian Optimal Interval Designs for Phase I
Clinical Trials. *Journal of the Royal Statistical Society: Series C*, 64,
507-523.
